No general trend for multiple dimensions of functional diversity in bird and mammal communities

### 2024-10-28 paper/GEB Submission/manuscript_GEB.Rmd
Minor abastract change

  Results: We found no overall temporal trends in functional diversity metrics. Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities. We found evidence of a temporal increase in fish consumption across all communities but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.

@@

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
csl: pnas.csl
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
  rticles::arxiv_article:
    keep_tex: true
#  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
keywords:
  - biodiversity change
  - functional traits
  - global change
  - time series

@@

To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species for abundance-based studies.
Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability.
`r colorize("This approach assumes that the likelihood of a species being present is equally likely through time and that the species abundance distribution is conserved.", "blue")` This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
Hereafter we refer to the *SES* as the corrected metric in comparison to the observed metric.
This approach will be less accurate for shorter time series, as we likely will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.

@@

'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

# Acknowledgments

We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits.

We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database.

>>>>

  Results: We found no overall temporal trends in functional diversity metrics. Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities. We found evidence of a temporal increase in fish consumption across all communities but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in bird foraging strata.

@@

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
csl: pnas.csl
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
  # rticles::arxiv_article:
  #   keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
keywords:
  - biodiversity change
  - functional traits
  - global change
  - time series

@@

To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species for abundance-based studies.
Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability.
`r colorize("This approach assumes the species pool is constant through time and that the species abundance distribution is conserved.", "blue")` This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
Hereafter we refer to the *SES* as the corrected metric in comparison to the observed metric.
This approach will be less accurate for shorter time series, as we likely will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.

@@

'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@
